New lower bounds for Schur and weak Schur numbers

### 2021-04-16 article/schur.tex
reorganisation de l'explication

\section{Schur numbers}

\qquad In this section, we use Rowley's constructions \cite{RowleyRamsey} in the context of Schur numbers. To improve lower bounds
for Ramsay's numbers, Rowley introduces some sum-free partitions verifying some additional properties which can be extended using a 
method which generalizes Abbott and Hanson's  construction \cite{AbbottHanson}. Rowley named these partitions "templates", and 
we will keep this name in the entire article. We find suitable templates and use them to provide new lower bounds for Schur numbers.

@@

\begin{definition}
We call SF-template of \(n\) colors and lenght \(p\) a partition of \( [\![1,p]\!]\) into \(n\) sum-free subsets \(A_1,
A_2, ..., A_n\) which verify :
	\[
	\forall i \in [\![1, n-1]\!], \forall (x,y) \in A_i^2, x+y > p
	\Longrightarrow x+y-p \notin A_i
	\]
	We note \(S^+(n)\) the maximal lenght for a SF-template of \(n\) colors.
\end{definition}

\begin{remark}
	Here, \(n\) is the "special" color: it has less constraints than the other colors. However, please note that \(n\) 
	is not necessarily the last color by order of appearance.
\end{remark}

\begin{remark}
	\(SF\)-templates include Abott and Hanson's construction \cite{AbbottHanson} as a special case.
\end{remark}

\begin{proposition}
	Let \(n \in [\![2, +\infty]\!]\). Then
	\[
	2S(n-1)+1 \leqslant S^+(n) \leqslant S(n)
	\]
\end{proposition}

\textsc{Proof :} The lower bound comes from Abott and Hanson's construction. The upper bound comes
from the fact that a SF-template of length \(p\) with \(n\) colors is also a partition of
\([\![1, p]\!]\) into \(n\) sum-free subsets.

@@

\qquad The main result on SF-templates follows. It allows us to improve lower bounds on Schur numbers by finding 
suitable SF-templates.

\begin{theorem}
	Let \((n,k), (p,q) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template of \(k+1\) colors and lenght \(p\),
	and a partition of \(n\) sum-free subsets of \([\![1,q]\!]\) then there exists a partition of \(n+k\) sum-free subsets
	of \([\![1,pq+m_{k+1}-1]\!]\). \(m_{k+1}\) is the minimum of the special subset in the \(SF\)-template.
\end{theorem}

@@

\begin{proposition}
Let \((k, p) \in \mathbb{N}^*)^2\) and let \(f\) be a colouring associated to a SF-template of length \(p\) with \(k+1\) colors. Let 
\(b \in \mathbb{N}\) (\(b = m_{k+1} - 1\) works) and assume there there exists a colouring \(g\) of \([\![1, b]\!]\) with \(k+1\) colors such that:
\begin{itemize}
	\item \(\forall (x, y) \in [\![1, p]\!]^2,(f(x) = f(y) \text{ and } (x + y) \mod p \leqslant b) 
	\implies g((x + y) \mod p) \neq f(x)\)
	\item \(\forall (x, y) \in [\![1, p]\!] \times  [\![1, b]\!], (f(x) = g(y) \text{ and } x + y \leqslant b) 
	\implies g(x + y) \neq f(x)\)
\end{itemize}
Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n))\), we have\\
\[ S(n+k) \geqslant S^+(k+1)S(n) + b\]
\end{proposition}

The idea lying beneath this theorem is similar to Abbott and Hanson's contruction \cite{AbbottHanson}. 
They extend vertically a sum-free partition, and horizontally an other sum-free partition. This way each "block" 
acts like a security zone for the other one. Here, the horizontal partition is no longer to the side of the 
vertical one, but it occupies the column of the special color of the \(SF\)-template, i.e the one without the 
extra condition. We give the following example for \(p = 9, q = 4, n = 2 ~\textrm{and}~ k = 2\). \\

This shows the inequality \(S(2+2) \geqslant S^+(3)S(2) + 4\). The special color is blue and is first used
to color 5, hence \(4 = m_{3} - 1\). The special color is blue.

\begin{center}
\[
\begin{array}{|*{9}{c|}}
	\multicolumn{9}{c}{\overset{S^+(3)}{\overbrace{\rule{6cm}{0pt}}}} \\
	\hline
	\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{blue} 5 & \cellcolor{blue} 6 & \cellcolor{red} 7 & \cellcolor{blue} 8 & \cellcolor{blue} 9 \\
	\hline
	\cellcolor{red} 10 & \cellcolor{green} 11 & \cellcolor{green} 12 & \cellcolor{red} 13 & \cellcolor{yellow} 14 & \cellcolor{yellow} 15 & \cellcolor{red} 16 & \cellcolor{yellow} 17 & \cellcolor{yellow} 18\\
	\hline
	\cellcolor{red} 19 & \cellcolor{green} 20 & \cellcolor{green} 21 & \cellcolor{red} 22 & \cellcolor{yellow} 23 & \cellcolor{yellow} 24 & \cellcolor{red} 25 & \cellcolor{yellow} 26 & \cellcolor{yellow} 27\\
	\hline
	\cellcolor{red} 28 & \cellcolor{green} 29 & \cellcolor{green} 30 & \cellcolor{red} 31 & \cellcolor{blue} 32 & \cellcolor{blue} 33 & \cellcolor{red} 34 & \cellcolor{blue} 35 & \cellcolor{blue} 36\\
	\hline
	\cellcolor{red} 37 & \cellcolor{green} 38 & \cellcolor{green} 39 & \cellcolor{red} 40 \\
	\cline{1-4}
\end{array}
\]
\end{center}

\begin{center}
\[
S^+(3) 
\begin{array}{|*{9}{c|}}
	\hline 
	\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{blue} 5 & \cellcolor{blue} 6 & \cellcolor{red} 7 & \cellcolor{blue} 8 & \cellcolor{blue} 9\\
	\hline
\end{array}  \]
\[
S(2) 
\begin{array}{|*{4}{c|}}
	\hline 
	\cellcolor{blue} 1 & \cellcolor{yellow} 2 & \cellcolor{yellow} 3 & \cellcolor{blue} 4 \\
	\hline
\end{array}
\]
\end{center}

To comparison, here is an usual Abbott and Hanson partition : \(S(2+2) \geqslant S(2)(2S(2)+1)+S(2) \). 
\begin{center}
\[
\begin{array}{|*{9}{c|}}
	\multicolumn{4}{c}{\overset{S(2)}{\overbrace{\rule{2.6cm}{0pt}}}} \\
	\hline
	\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{blue} 5 & \cellcolor{blue} 6 & \cellcolor{blue} 7 & \cellcolor{blue} 8 & \cellcolor{blue} 9\\
	\hline
	\cellcolor{red} 10 & \cellcolor{green} 11 & \cellcolor{green} 12 & \cellcolor{red} 13 & \cellcolor{yellow} 14 & \cellcolor{yellow} 15 & \cellcolor{yellow} 16 & \cellcolor{yellow} 17 & \cellcolor{yellow} 18\\
	\hline
	\cellcolor{red} 19 & \cellcolor{green} 20 & \cellcolor{green} 21 & \cellcolor{red} 22 & \cellcolor{yellow} 23 & \cellcolor{yellow} 24 & \cellcolor{yellow} 25 & \cellcolor{yellow} 26 & \cellcolor{yellow} 27\\
	\hline
	\cellcolor{red} 28 & \cellcolor{green} 29 & \cellcolor{green} 30 & \cellcolor{red} 31 & \cellcolor{blue} 32 & \cellcolor{blue} 33 & \cellcolor{blue} 34 & \cellcolor{blue} 35 & \cellcolor{blue} 36 \\
	\hline
	\cellcolor{red} 37 & \cellcolor{green} 38 & \cellcolor{green} 39 & \cellcolor{red} 40 \\
	\cline{1-4}
\end{array}
\]
\end{center}

\textsc{Proof of theorem:} We denote by \(f\) the colouring associated to the \(SF\)-template of lenght \(p\)
and \(g\) the one associated to the sum-free partition of \([\![1,q]\!]\).

@@

@@

The previous inequalities give new lower bounds for \(S(n)\) for
\( n \geqslant 9 \). We compute the lower
bounds for \( n \in [\![8,15]\!] \) using the four different inequalities, please notice that the best values for \( n
=8\) and \(n = 13\) were obtained thanks to the first one, found by Rowley. The best lower bounds are highlighted.\\
\\
\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	\(n\) & 8 & 9 & 10 & 11 \\
	\hline
	\(33S(n-3) + 6 \) & \cellcolor{yellow} 5286 & 17694 & 55446 & 174444\\
	\hline
	\(111S(n-4) + 43 \) & 4927 & \cellcolor{yellow} 17803 & 59539 & 186523\\
	\hline
	\(380S(n-5) + 148 \) & 5088 & 16868 & \cellcolor{yellow} 60948 & \cellcolor{yellow} 203828 \\
	\hline
	\(1140S(n-6) + 528 \) & 5088 & 15348 & 50688 & 182928\\
	\hline
	\hline
	\(n\) & 12 & 13 & 14 & 15 \\
	\hline
	\(33S(n-3) + 6 \) & 587505 & \cellcolor{yellow} 2011290 & 6726330 & 21072090\\
	\hline
	\(111S(n-4) + 43 \) & 586789 & 1976176 & 6765271 & 22624951 \\
	\hline
\(380S(n-5) + 148 \) & \cellcolor{yellow} 638548 & 2008828 & \cellcolor{yellow} 6765288 & \cellcolor{yellow} 23160388 \\\hline
	\(1140S(n-6) + 528 \) & 611568 & 1915728 & 6026568 & 20295948 \\
	\hline
\end{tabular}
\end{center}
Except for 8, 9 and 13, the best lower bounds are obtained thanks to
the third inequality \( S(n+5) \geqslant 380S(n) + 148\). The table
doesn't go any further, but the same inequality allows to improve the
lower bounds for every \( n \geqslant 15 \).

>>>>

\section{Schur numbers}
\label{Schur}

\qquad In this section, we use Rowley's template-based constructions \cite{RowleyRamsey} in the context of Schur 
numbers. To improve lower bounds for Ramsay's numbers, Rowley introduces special sum-free partitions verifying 
some additional properties which can be extended using a method generalizing Abbott and Hanson's construction 
\cite{AbbottHanson}. Rowley named these partitions "templates", and we keep this name in the entire article. We 
then find new templates and use them to provide new lower bounds for Schur numbers.

@@

\begin{definition}
A SF-template with length \(p\) and \(n\) colors is defined as a partition of \( [\![1,p]\!]\) into \(n\) sum-free subsets 
\(A_1, A_2, ..., A_n\) verifying:
\[
\forall i \in [\![1, n-1]\!], \forall (x,y) \in A_i^2, x+y > p
\Longrightarrow x+y-p \notin A_i
\]
\end{definition}

\begin{remark}
	Here \(n\) is the "special" color: it has less constraints than the other colors. However, please note that \(n\) 
	is not necessarily the last color by order of appearance.
\end{remark}

\begin{remark}
	\(SF\)-templates include Abbott and Hanson's construction \cite{AbbottHanson} as a special case.
\end{remark}

\begin{proposition}
	Let \(n \in [\![2, +\infty[\![\). We define \(S^+(n)\) as the maximal lenght of a SF-template with \(n\) colors. 
	\(S^+(n)\) is well defined and verifies:
	\[
	2S(n-1)+1 \leqslant S^+(n) \leqslant S(n)
	\]
\end{proposition}

\textsc{Proof :} The lower bound comes from Abbott and Hanson's construction. The upper bound comes from the 
fact that a SF-template with length \(p\) and \(n\) colors is also a partition of \([\![1, p]\!]\) into \(n\) sum-free subsets.

@@

\qquad Here we state the main result on SF-templates stated by Rowley in the context of Ramsey numbers. It consists 
in the explicit construction of a sum-free partition using a SF-template and a sum-free partition.

\begin{theorem}
	Let \((n,k), (p,q) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template with lenght \(p\) and \(k+1\) colors 
	and a partition of \([\![1,q]\!]\) into \(n\) sum-free subsets then there exists a partition of \([\![1,pq+m_{k+1}
	-1]\!]\) into \(n+k\) sum-free subsets. \(m_{k+1}\) is the minimum of the special subset in the \(SF\)-template.
\end{theorem}

@@

The idea lying beneath this theorem is similar to Abbott and Hanson's contruction \cite{AbbottHanson}. They vertically 
extend a sum-free partition by repeating it and they use an other sum-free partition to color the other half according 
to the line number. This way the "blocks" act as safe areas for each other. We give here an example for \(p = 9\), 
\(q = 4\), \(n = 2\) and \(k = 2\) showing that \(S(2 + 2) \geqslant S(2) (2 S(2) + 1) + S(2)\), both with Abbott 
and Hanson's construction and with a SF-template which is not of this type. In both cases, the special color is grey.

\begin{center}
\textbf{Abbott and Hanson's construction}
\end{center}

\[
\begin{array}{|*{9}{c|}}
	\hline
	\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{cyan} 5 & 
	\cellcolor{cyan} 6 & \cellcolor{cyan} 7 & \cellcolor{cyan} 8 & \cellcolor{cyan} 9\\
	\hline
	\cellcolor{red} 10 & \cellcolor{green} 11 & \cellcolor{green} 12 & \cellcolor{red} 13 & \cellcolor{yellow} 14 & 
	\cellcolor{yellow} 15 & \cellcolor{yellow} 16 & \cellcolor{yellow} 17 & \cellcolor{yellow} 18\\
	\hline
	\cellcolor{red} 19 & \cellcolor{green} 20 & \cellcolor{green} 21 & \cellcolor{red} 22 & \cellcolor{yellow} 23 & 
	\cellcolor{yellow} 24 & \cellcolor{yellow} 25 & \cellcolor{yellow} 26 & \cellcolor{yellow} 27\\
	\hline
	\cellcolor{red} 28 & \cellcolor{green} 29 & \cellcolor{green} 30 & \cellcolor{red} 31 & \cellcolor{cyan} 32 & 
	\cellcolor{cyan} 33 & \cellcolor{cyan} 34 & \cellcolor{cyan} 35 & \cellcolor{cyan} 36 \\
	\hline
	\cellcolor{red} 37 & \cellcolor{green} 38 & \cellcolor{green} 39 & \cellcolor{red} 40 \\
	\cline{1-4}
\end{array}
\]

\begin{center}
\begin{tabular}{c c}
	\textbf{Corresponding SF-template} & \textbf{Corresponding sum-free partition} \\
	\begin{tabular}{|*{9}{c|}}
		\hline 
		\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{gray!40} 5 & 
		\cellcolor{gray!40} 6 & \cellcolor{gray!40} 7 & \cellcolor{gray!40} 8 & \cellcolor{gray!40} 9 \\
		\hline
	\end{tabular} &
	\begin{tabular}{|*{4}{c|}}
		\hline 
		\cellcolor{cyan} 1 & \cellcolor{yellow} 2 & \cellcolor{yellow} 3 & \cellcolor{cyan} 4 \\
		\hline
	\end{tabular}
\end{tabular}
\end{center}

In the general construction with SF-templates, the special color no longer necessarily contains consecutive 
numbers. However, the special color is still replaced by the colors of the sum-free partition according to the 
line number and the other colors are still vertically extended.

\begin{center}
\textbf{SF-template construction}
\end{center}

\[
\begin{array}{|*{9}{c|}}
	\hline
	\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{cyan} 5 & 
	\cellcolor{cyan} 6 & \cellcolor{red} 7 & \cellcolor{cyan} 8 & \cellcolor{cyan} 9\\
	\hline
	\cellcolor{red} 10 & \cellcolor{green} 11 & \cellcolor{green} 12 & \cellcolor{red} 13 & \cellcolor{yellow} 14 & 
	\cellcolor{yellow} 15 & \cellcolor{red} 16 & \cellcolor{yellow} 17 & \cellcolor{yellow} 18\\
	\hline
	\cellcolor{red} 19 & \cellcolor{green} 20 & \cellcolor{green} 21 & \cellcolor{red} 22 & \cellcolor{yellow} 23 & 
	\cellcolor{yellow} 24 & \cellcolor{red} 25 & \cellcolor{yellow} 26 & \cellcolor{yellow} 27\\
	\hline
	\cellcolor{red} 28 & \cellcolor{green} 29 & \cellcolor{green} 30 & \cellcolor{red} 31 & \cellcolor{cyan} 32 & 
	\cellcolor{cyan} 33 & \cellcolor{red} 34 & \cellcolor{cyan} 35 & \cellcolor{cyan} 36 \\
	\hline
	\cellcolor{red} 37 & \cellcolor{green} 38 & \cellcolor{green} 39 & \cellcolor{red} 40 \\
	\cline{1-4}
\end{array}
\]

\begin{center}
\begin{tabular}{c c}
	\textbf{Corresponding SF-template} & \textbf{Corresponding sum-free partition} \\
	\begin{tabular}{|*{9}{c|}}
		\hline 
		\cellcolor{red} 1 & \cellcolor{green} 2 & \cellcolor{green} 3 & \cellcolor{red} 4 & \cellcolor{gray!40} 5 & 
		\cellcolor{gray!40} 6 & \cellcolor{red} 7 & \cellcolor{gray!40} 8 & \cellcolor{gray!40} 9 \\
		\hline
	\end{tabular} &
	\begin{tabular}{|*{4}{c|}}
		\hline 
		\cellcolor{cyan} 1 & \cellcolor{yellow} 2 & \cellcolor{yellow} 3 & \cellcolor{cyan} 4 \\
		\hline
	\end{tabular}
\end{tabular}
\end{center}

We know proceed to prove the above mentioned theorem.\\

\textsc{Proof:} We denote by \(f\) the colouring associated to the \(SF\)-template of lenght \(p\)
and \(g\) the one associated to the sum-free partition of \([\![1,q]\!]\).

@@

\begin{proposition}
Let \((k, p) \in \mathbb{N}^*)^2\) and let \(f\) be a colouring associated to a SF-template of length \(p\) with \(k+1\) 
colors. Let \(b \in \mathbb{N}\) (\(b = m_{k+1} - 1\) works) and assume there there exists a colouring \(g\) of 
\([\![1, b]\!]\) with \(k+1\) colors such that:

\begin{itemize}
	\item \(\forall (x, y) \in [\![1, p]\!]^2,(f(x) = f(y) \text{ and } (x + y) \mod p \leqslant b) 
	\implies g((x + y) \mod p) \neq f(x)\)
	\item \(\forall (x, y) \in [\![1, p]\!] \times  [\![1, b]\!], (f(x) = g(y) \text{ and } x + y \leqslant b) 
	\implies g(x + y) \neq f(x)\)
\end{itemize}

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n))\), we have
\[ S(n+k) \geqslant S^+(k+1)S(n) + b\]
\end{proposition}

@@

The previous inequalities give new lower bounds for \(S(n)\) for
\( n \geqslant 9 \). We compute the lower
bounds for \( n \in [\![8,15]\!] \) using the four different inequalities, please notice that the best values for \( n
=8\) and \(n = 13\) were obtained thanks to the first one, found by Rowley. The best lower bounds are highlighted.

\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	\(n\) & 8 & 9 & 10 & 11 \\
	\hline
	\(33S(n-3) + 6 \) & \cellcolor{yellow} 5286 & 17694 & 55446 & 174444\\
	\hline
	\(111S(n-4) + 43 \) & 4927 & \cellcolor{yellow} 17803 & 59539 & 186523\\
	\hline
	\(380S(n-5) + 148 \) & 5088 & 16868 & \cellcolor{yellow} 60948 & \cellcolor{yellow} 203828 \\
	\hline
	\(1140S(n-6) + 528 \) & 5088 & 15348 & 50688 & 182928\\
	\hline
	\hline
	\(n\) & 12 & 13 & 14 & 15 \\
	\hline
	\(33S(n-3) + 6 \) & 587505 & \cellcolor{yellow} 2011290 & 6726330 & 21072090\\
	\hline
	\(111S(n-4) + 43 \) & 586789 & 1976176 & 6765271 & 22624951 \\
	\hline
\(380S(n-5) + 148 \) & \cellcolor{yellow} 638548 & 2008828 & \cellcolor{yellow} 6765288 & \cellcolor{yellow} 23160388 \\\hline
	\(1140S(n-6) + 528 \) & 611568 & 1915728 & 6026568 & 20295948 \\
	\hline
\end{tabular}
\end{center}
Except for 8, 9 and 13, the best lower bounds are obtained thanks to
the third inequality \( S(n+5) \geqslant 380S(n) + 148\). The table
doesn't go any further, but the same inequality allows to improve the
lower bounds for every \( n \geqslant 15 \).
